---
title: "Pandemic Flu Simulation And The Impact Of Ultraviolet Light On Reproduction Ratio"
author: "Nathan Kurtz-Enko"
format: pdf
toc: true
echo: false
warning: false
message: false
bibliography: references.bib
---

# Abstract

Influenza and similar viruses represent a threat to the health populations across the globe. It is important to study the behavior of these viruses and the impact of mitigation strategies. Compartmental and stochastic models have been used to show that mitigation strategies like vaccination, prophylaxis antiviral agents, and travel restrictions can have some impact on or fully contain the spread of influenza. While some strategies have been more widely studied, others, such as ultraviolet (UV) radiation treament of air and surfaces, have received less attention. We examined the number of days of UV treatment required to contain pandemic influenza in the United States of America with an initial reproduction ratio of $R_0 = 1.8$. During a 100 day simulation, beginning once 5% of the overall population has been infected, we determined that approximately 30 days of UV treatment would reduce the reproduction ratio of influenza to $R_0 \approx 1$, meaning that the spread of the virus is no longer in an outbreak status.

# Introduction

Understanding the spread of infectious diseases, like the flu, is critical for advising policy making with respect to public health and wellness. For example, studying 
the behavior of virus transmission and recovery rates allows health professionals and policy-makers to define the quantity of antiviral agents that should be readily available, regions and demographics that should be prioritized for vaccination, and other strategies intended to prevent a pandemic. Furthermore, being able to communicate the results of an infectious disease model/simulation is very important, as this will facilitate the adoption of containment measures.

Many researchers studying the spread of infectious diseases explicitly cite the 1918 influenza outbreak as a motivating factor for their work [@berkessel2021social; @germann2006mitigation; @johnson2002spanishflu; @valle2012behavior]. In the case of @johnson2002spanishflu and @patterson1991geography, the objective was to look back through previous records and data to verify morbitity of the 1918 pandemic. Most recently, @johnson2002spanishflu estimated the death toll from the influenza was on the order of 50 million people.

Clearly, the capacity for large-scale damage from infectious disease outbreaks is a very real threat. Hence, understanding how these viruses spread and the impact that different mitigation strategies have on the rate of spread and severity of illness is critical.

## Literature Review

In the 1980s, Russian scientists first began to implement models for estimating the spread influenza and other viruses [@longini1986global]. Eventually, the work made its way into academia across the globe. These early models were deterministic, and they described the movement of viruses using differential equations, and partitioned populations into several subgroups. These are referred to as SIR (i.e., susceptible, infected, and recovered groups), SEIR (i.e., susceptible, exposed, infected, and recovered groups), and other derivatives.

Initial reactions to these determinsitic models by academics were skeptical for a few reasons. First, geographic regions being studied were relatively isolated, hence the ability to apply these models for global study was uncertain [@longini1986global]. Second, for the time, the models were computationally expensive, meaning it was impractical to use them. So, academics remained skeptical of these models' utility. However, this is less of an issue today because high-performance computing is readily available.

While relatively old, determinstic models still have utility. @valle2012behavior used an SIR model to show that an individual's behavior will likely change in response to an outbreak; some individuals will adopt new behavior to protect themselves, thereby influencing the probability of transmission per contact. 

More recently, stochastic models have become prevalent because they are able to capture more complex spreading patterns. @colizza2007global, @cooper2006delaying, @ferguson2005containment, @germann2006mitigation, @samsuzzoha2013parameter, and @tan2021stochastic all used stochastic processes to model the spread of influenza or other infectious diseases, and analyze different mitigation strategies. @colizza2007global found that a large-scale application of prophylaxis antiviral agents could successfully contain an H5N1 influenza pandemic given a viral reproduction ratio of 1.9 or lower. @cooper2006delaying found that significant air travel restrictions could delay the spread of influenza but not contain an epidemic altogether. The findings of @germann2006mitigation determined that rapid and preferential vaccination of children could contain an outbreak when the reproduction ratio of influenza is at or below 1.9.

Others have explored the relationship between geographic variables (e.g., altitude and solar radiation) and incidence rates for COVID-19, as well as severity of illness [@arias2021altitude; @stephens2021altitude]. @arias2021altitude concluded that above 1000 meters above sea level, incidence rates and severity for COVID-19 decline. Through analysis of public geographic and COVID-19 data, @stephens2021altitude determined that for every increase of 495 meters in elevation above LA county, and for areas of equivalent population density, infection rates were 12.82%, 12.01%, and 11.72% respectively. However, mortality rates were not statistically significantly different between high and low elevation areas. @stephens2021altitude note that other environmental variables such as increased solar radiation, larger swings of high/low temperatures throughout the day, and long-term exposure to altitude hypoxia could explain these observed differences.

Since influenza and similar viruses spread via aerosols, @sagripanti2007solar proposed solar radiation as a contributing factor to the seasonality of influenza. Specifically, the authors calculated the expected inactivation of influenza by UV radiation in several cities and different times of the year. They estimated that a full day of sunlight exposure will reduce influenza by 99% in regions at a similar latitude to Mexico City, Mexico, and by 90% in regions at a similar latitude to Miami, Florida. @jensen1964inactivation found that aerosol influenza A, when exposed to 0.03 watt-minutes per square foot (i.e., approximately 19.41286 Joules per square meter) of UV light, had an inactivation of greater than 99.9% at a flow rate 100 cubic feet per minute, and approximately 99.86% inactivation at 200 cubic feet per minute. 

Therefore, higher altitude and higher solar radiation could theoretically influence the spread of influenza. The purpose of this paper is to simulate pandemic flu spread and identify critical thresholds for some of these covariates in order to minimize the risk of a severe outbreak.

## Modeling Theory

As previously mentioned, there are two very general approaches to simulating pandemic virus spread. Here, we present the SIR model [@debarre_sir; @wb_deterministic] because it is relatively simple, and establishes a foundation for considering other approaches.

The SIR model categorizes a population into three groups: susceptible, infected, and recovered. The rate of change of each of these groups is defined using differential equations, and we can account for mortality and birth rates in their definitions as well.

$$
\begin{aligned}
\frac{dS}{dt} =& \delta - \frac{\beta \times S \times I}{N} - \mu \times S \\
\frac{dI}{dt} =& \frac{\beta \times S \times I}{N} - (\gamma + \mu) \times I \\
\frac{dR}{dt} =& \gamma \times I - \mu \times R
\end{aligned}
$$

Here, $S$, $I$, $R$ stand for the susceptible, infected, and recovered populations respectively. The total population is denoted by $N$, and the rates of transmission, recovery, mortality, and birth are denoted by $\beta$, $\gamma$, $\mu$, and $\delta$.

The outbreak status is described by the reproduction ratio, $R_0$, which is the ratio between the transmission rate and recovery rate. When $R_0 = \beta / \gamma > 1$, more people are becoming infected than are recovering, hence the system is in a state of outbreak. And, supposing that there is some percentage of the population that is already vaccinated (i.e., already in the recovered group), this ratio becomes:

$$
R_0 = (\beta \times S(0))/(\gamma \times N)
$$

Here, $S(0)$ is the number of people in the susceptible group at time $t=0$.

These facts make it easy to define different transmission, recovery, and vaccination target rates to avoid a significant outbreak. Hence, the SIR is a simple model, but effective at clearly communicating actionable results to policy makers and authority figures.

# Method

In this analysis, we used Epiverse-TRACE [@epidemics2025composable; @epiparameter2025classes ;@finalsize2025calculate], Tidyverse [@dplyr2025manipulation; @purrr2025functional; @tidyr2025tidy; @readr2024rectangular; @ggplot22016elegant], and other tools [@cdcfluview2022retrieve; @tidycensus2025load; @knitr2025general] in the R programming language. The objective was to use contact and population estimates to implement an SEIR mode and simulate the spread of influenza in the United States of America.

```{r}
# Double check project state
# renv::status()
# renv::snapshot()
# renv::restore()
# Install packages
# renv::install(c("dplyr", "tidyr", "purrr", "readr", "ggplot2"))
# renv::install(c("finalsize", "epiparameter"))
# renv::install("epidemics", repos = 'https://epiverse-trace.r-universe.dev')
# renv::install(c("tidycensus", "knitr"))

# Tidyverse tools
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
# Epiverse-TRACE tools
library(finalsize)
library(epiparameter)
library(epidemics)
# Other tools
library(knitr)
library(tidycensus)
```

## Data

We used contact data collected by @breen2022novel for different age groups across the United States of America. Additionally, we accessed U.S. Census results from 2022 via @tidycensus2025load to estimate populations for each of the age groups defined by @breen2022novel. Census population estimates for contact data age groups were not provided by @tidycensus2025load, however, we aggregated the state-wide counts for all ages, ranging from one to 100, into bins defined in the contact data. The data are displayed in @tbl-contacts and @tbl-demography.

```{r}
# Load contact & population data
usa_files <- list.files(path = "data/predictions", pattern = ".rds$", full.names = TRUE) 
usa_data <- map(usa_files, read_rds, .progress = TRUE) 
usa_contacts <- bind_rows(map(usa_data, \(x) {x[[2]]}, .progress = TRUE))
usa_population <- get_estimates(
  geography = "state",
  product = "characteristics",
  variables = "POPESTIMATE",
  breakdown = "AGE",
  vintage = 2022
)

# Build a contact matrix
usa_matrix <- ungroup(usa_contacts) |>
  reframe(
    across(all_of(c("contacts", "lower", "upper")), sum), 
    .by = c(".ego_age", ".alter_age")
  ) |>
  arrange(.ego_age, .alter_age) |>
  pull(contacts) |>
  matrix(nrow = 7, ncol = 7)
# .ego_age is the age group and represents column names
# .alter_age is the contact age group and represents row names
rownames(usa_matrix) <- sort(unique(usa_contacts$`.alter_age`))
colnames(usa_matrix) <- sort(unique(usa_contacts$`.ego_age`))

# Build demography vector
usa_population$age_group <- cut(
  usa_population$AGE, 
  breaks = c(0, 18, 25, 35, 45, 55, 65, 101), 
  right = FALSE
)
usa_demography <- usa_population |>
  reframe(value = sum(value), .by = "age_group") |>
  arrange(age_group)
usa_demography$age_group <- rownames(usa_matrix)
usa_demography <- pull(usa_demography, value) |>
  set_names(usa_demography$age_group)
```

```{r}
#| label: tbl-contacts
#| tbl-cap: "USA National Contact Matrix"

kable(usa_matrix)
```

```{r}
#| label: tbl-demography
#| tbl-cap: "USA National Population Estimates"

tibble(`Age Group` = names(usa_demography), Population = usa_demography) |>
  kable()
```

## Initial Conditions

There are two approaches to defining initial conditions (e.g., number of members in each model sub-group at initialization, influenza spread parameters, etc.) for the model: define some reasonable values or derive values from current/historic flu data. For convenience, we used the former approach and use values for rates discussed in literature. 

@royanikbakht2018estimation provided multiple estimates for the basic reproduction ratio of influenza in the United states given 2017-2018 data. Specifically, using Maximum Likelhood Estimation, they estimated $R_0 = 1.8$ with a 95% confidence interval of $[1.78, 1.8]$. While this estimate is approaching seven years old, it should be sufficient for the purpose of this paper.

@virlogeux2015estimating provided an estimate for the incubation period of influenza. Specifically, they noted that incubation period follows a Weibull distribution with $\text{shape} = 2.101$ and $\text{scale} = 3.839$. We used the expected value of the incubation period probability distribution.

$$
E[\text{Incubation Period}] = \text{scale} \times \Gamma \left( 1 + \frac{1}{\text{shape}} \right) = 3.400167 \text{ days}
$$ 

@tang2020estimation noted that the inverse incubation period is equivalent to the infection rate, therefore $\alpha = 0.2941032$.

According to @cdc2024influenza, the infectious period of influenza is about five to seven days typically. For the purpose of this paper, we assumed a worst case scenario and used an infectious period of seven days. @turincini2025impact stated that the inverse of the infectious period is equivalent to the recovery rate, therefore $\gamma = 0.1428571$.

The final parameters are collected and displayed in @tbl-params.

```{r}
incubation_params <- epiparameter_db(
  disease = "influenza",
  epi_name = "incubation period",
  single_epiparameter = TRUE
)
incubation_period <- incubation_params$summary_stats$mean
infectious_period <- 7
R_0 <- 1.8 # Reproduction ratio
alpha <- 1 / incubation_period # Infection rate
gamma <- 1 / infectious_period # Recovery rate
beta <- R_0 * gamma # Transmission rate
```

```{r}
#| label: tbl-params
#| tbl-cap: "Parameters for SEIRV Model"

kable(
  tibble(
    Name = c("Reproduction Ratio", "Infection Rate", "Recovery Rate", "Transmission Rate"),
    Symbol = c("$R_0$", "$\\alpha$", "$\\gamma$", "$\\beta$"),
    Value = c(R_0, alpha, gamma, beta)
  )
)
```

In order for the flu to progress through a population, there must be some members in the infected group at the start of the simulation. We assumed that 5% of the population is already infected. Furthermore, we assumed that this proportion of infected individuals was consistent across all age groups.  

```{r}
# Define initial conditions
n <- sum(usa_demography)
initial_i <- 0.05
initial_conditions <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)
initial_conditions <- rownames(usa_matrix) |>
  map(\(x) initial_conditions) |>
  bind_rows() |>
  as.matrix()
rownames(initial_conditions) <- rownames(usa_matrix)

# Create the population object
population_data <- population(
  name = "USA",
  contact_matrix = usa_matrix,
  demography_vector = usa_demography,
  initial_conditions = initial_conditions
)
```

## Mitigation Strategies

As noted by @jensen1964inactivation and @sagripanti2007solar, solar radiation and ultraviolet light can effectively inactivate influenza. One mitigation strategy, in addition to encouraged vaccination and applications of antiviral agents, could be UV radiation dosing, air sanitization in buildings, and surface sanitization in other settings.

Since @jensen1964inactivation determined that approximately 19 Joules per square meter of UV light results in over 99.9% viral inactivation, we used those figures in our intervention measure. The supposition we examined is that this strategy could be used as a reactionary measure to an emergent epidemic, rather than a preventative one, and our objective was to define the number of days of this treatment required to curtail a pandemic. 

```{r}
uv_radiation <- map(
  2:100,
  \(x) {
    intervention(
      type = "rate",
      time_begin = 1, # Days
      time_end = x, # Also days
      reduction = 0.99
    )
  }
)
```

## Simulation Results

For this simulation, we used a SEIR model, and we compared a base output, in which there are no intervention measures taken, with alternative outcomes in which a UV exposure intervention has been made.

```{r}
base_output <- model_default(
  population = population_data,
  transmission_rate = beta,
  infectiousness_rate = alpha,
  recovery_rate = gamma,
  time_end = 100, # Days 
  increment = 1.0 # Also days
)
intervention_output <- map(
  uv_radiation,
  \(x) {
    model_default(
      population = population_data,
      transmission_rate = beta,
      infectiousness_rate = alpha,
      recovery_rate = gamma,
      intervention = list(transmission_rate = x),
      time_end = 100, # Days 
      increment = 1.0 # Also days
    )
  }
)
```

@andreasen2011finalsize defines the relationship between basic reproduction ratio and the final size of the epidemic. We used a simplified implementation of this work to estimate the reproduction ratio after the intervention methods were applied. Here, $S_0$ is the population of susceptible individuals at time $t=0$ and $S_{100}$ is the population of susceptible individuals at time $t=100$ (i.e., the end of the simulation). 

$$
R_0 = ln \left( \frac{S_{100}}{S_0} \right) / \left( 1 - \frac{S_{100}}{S_0} \right) 
$$

We observed that after about 30 days of 19 Joules per square meter of UV radiation exposure, the reduction in the reproduction ratio reached steady state, indicating there is less benefit to additional application of UV light. This is presented in @fig-interventionratios. The simulation results of our base case (with no intervention applied) and the 30 days of UV treatment are presented in @fig-baseoutput and @fig-interventionoutput, and show a clear difference in the spread of influenza, with the former visualization depicting a well controlled virus.

```{r}
#| label: fig-interventionratios
#| fig-cap: "Influenza simulation resulting reproduction ratios spanning various intervention levels of applied UV radiation exposure."

est_reproduction_ratio <- function(x) {
  S <- reframe(x, value = sum(value), .by = c("compartment", "time")) |>
    filter(compartment == "susceptible")
  S_0 <- slice_min(S, order_by = time) |>
    pull(value)
  S_inf <- slice_max(S, order_by = time) |>
    pull(value)
  frac <- S_inf / S_0
  R0 <- -log(frac) / (1 - frac)
  return(R0)
}
reproduction_ratios <- tibble(
  days = 2:100,
  R_0 = map_vec(
    intervention_output,
    est_reproduction_ratio
  )
)
ggplot(data = reproduction_ratios) +
  geom_point(mapping = aes(x = days, y = R_0)) +
  geom_line(mapping = aes(x = days, y = R_0)) +
  theme_minimal() +
  labs(x = "Days of UV Exposure", y = "Reproduction Ratio")
```

```{r}
#| label: fig-baseoutput
#| fig-cap: "Influenza simulation output results without any intervention and spanning 100 days."

vis_infection <- function(x) {
  x <- filter(x, compartment != "vaccinated") |>
    reframe(value = sum(value), .by = c("time", "compartment"))
  plt <- ggplot(data = x) +
    geom_point(mapping = aes(x = time, y = value, color = compartment)) +
    theme_minimal() +
    scale_color_viridis_d() +
    labs(x = "Time (days)", y = "Population", color = "Compartment")
  return(plt)
}
vis_infection(base_output)
```

```{r}
#| label: fig-interventionoutput
#| fig-cap: "Influenza simulation output results with UV radiation intervention spanning day two until day 30 of the 100 day simulation."

vis_infection(intervention_output[[29]])
```

# Discussion

Assuming that the initial reproduction ratio, transmission rate, infection rate, and recovery rate of an influenza outbreak are respectively approximately `{r} R_0`, `{r} beta`, `{r} alpha`, and `{r} gamma`, then the results of this simulation suggest that an application of 19 Joules per square meter of UV exposure to air and surfaces that populations regularly interact with for 30 days after 5% of the population is infected should reduce the reproduction ratio to `{r} est_reproduction_ratio(intervention_output[[29]])`. This suggests that the mitigation strategy of using UV radiation to sanitize air/surfaces could significantly contribute to containing an influenza pandemic.

However, due to time constraints, this paper does not account for heterogeneity in age groups and contacts when estimating the reproduction ratio after mitigation strategies are applied. Furthermore, these results have not been validated against real world case data. Therefore, additional work to conclusively state the impact of UV radiation as a mitigation strategy should be performed. Finally, the practicality of implementing UV radiation as a mitigation step was not discussed. 

In future work, combined models of epidemiology, climatology, and public health could be used to capture and evaluate more complex interactions between age groups, the environment, and the impact of mitigation measures as well. Additionally, directly modeling an observed epidemic would give more credence to any conclusions about mitigation steps. And, discussing the methods for implementing mitigations steps would provide more support for conclusions.

So, while not definitive, this simulation does generally suggest that UV light can be used to mitigate/contain an influenza pandemic. 

# References